Building the Million Transcriptome Ocular Meta-Atlas

### 2020-11-16 manuscript_00.Rmd
fixes

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

```{r figBenchmark, fig.width=13, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d. Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')

source('figs_and_tables/fig_umap_benchmarking.R')
plot_grid(
  plot_grid(
    umap_plot_maker(umapCombat), 
    umap_plot_maker(umapNone, color_against = 'study_accession'),
    rel_widths = c(1.3, 1),
    labels = c('a','b')), 
  plot_grid(
    zscore_sum_all_methods, 
    zscore_droplet_scVI_optimize,
    labels = c('c','d')),
  nrow = 2,
  rel_heights = c(1, 0.4)
)
```
## scVI has highest integration performance

@@

```{r figXGboost_umap, fig.width=8, fig.height=13, out.width=600, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')

@@

input <- list()
input[['meta_column']] <- 'organism'
input[['pt_size_meta']] <- 1
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
#input[['meta_filter_cat']] <- 'CellType_predict'
#input[['meta_filter_on']] <- 'Bipolar Cells'
org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)
plot_grid(ct_alluvial, 
          plot_grid(ctp$plot, 
                    ctp$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
                    org$plot + facet_wrap(~organism) + theme(legend.position = "none"), 
                    ncol = 1,
                    labels = c('b', 'c', 'd'),
                    rel_heights = c(1,1,0.5)), 
          rel_widths = c(0.2,1), 
          labels = c('a'))

@@

## ML cell type labels allow for unprecedented biological in cell types

## The scVI-based scEiaD UMAP projection blends batches and separates cell types
The 2D UMAP projection of the scVI-calculated batch corrected 8 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types. We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization. The photoreceptors, cones and rods, of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The amacrine cells ....say more about cell type in space

## Massive pseudbulk differential testing leverages high biological diversity

@@

@@

benchmarked with:
- LISI
- ARI
- Silhouette
- PCA variance explained before/after correction

## Clustering
Louvain-jaccard as implemented by Seurat

Leiden by ? in anndata

@@

```{r supTab_AUCxgboost, fig.width=14,  fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'TEXT')

@@

```{r supTab_CTmislabels, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}

@@

#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(predictions %>%  
           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
           select(-cell_type_id) %>% 
           rename(PredCellType = CellType) %>% 
           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
           filter(!is.na(TrueCellType)) %>% 
           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
           group_by(PredCellType, TrueCellType) %>% 
           summarise(Count = n()) %>% 
           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
           filter(Ratio > 0.05) %>% 
           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
           flextable() %>% 
           width(width = 3), 
         size = 8, 
         part = 'all')
```

>>>>

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
library(grid)
library(ComplexHeatmap)
library(viridis)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

```{r figBenchmark, fig.width=8, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d. Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')

source('figs_and_tables/fig_umap_benchmarking.R')
plot_grid(
  plot_grid(
    umap_plot_maker(umapCombat), 
    umap_plot_maker(umapNone, color_against = 'study_accession'),
    rel_widths = c(1.3, 1),
    labels = c('a','b')), 
  plot_grid(
    zscore_sum_all_methods, 
    zscore_droplet_scVI_optimize,
    labels = c('c','d')),
  nrow = 2,
  rel_heights = c(1, 0.4)
)
```

## scVI has highest integration performance

@@

```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')

@@

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_meta']] <- 3
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
#input[['meta_filter_cat']] <- 'CellType_predict'
#input[['meta_filter_on']] <- 'Bipolar Cells'
ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

input <- list()
input[['meta_column']] <- 'organism'
input[['pt_size_meta']] <- 1
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
#input[['meta_filter_cat']] <- 'CellType_predict'
#input[['meta_filter_on']] <- 'Bipolar Cells'
org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)
plot_grid(grid.grabExpr(draw(ct_confusion)), 
          ctp$plot + theme(legend.position = "none"), 
          ctp2$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
          org$plot + facet_wrap(~organism) + theme(legend.position = "none"), 
          ncol = 2,
          labels = c('a', 'b', 'c', 'd'),
          rel_heights = c(1,0.8))

@@

## ML cell type labels allow for high diversity in studies for each cell type

## The scVI-based scEiaD UMAP projection blends batches and separates cell types
The 2D UMAP projection of the scVI-calculated batch corrected 8 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types. We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization. The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The amacrine cells ....say more about cell type in space

## Pseudobulk differential testing leverages high biological diversity

Discuss

@@

V. briefly discuss

@@

benchmarked with:
- LISI
- ARI
- NMI
- Silhouette
- PCA variance explained before/after correction

## Clustering
Louvain-jaccard as implemented by Seurat and leiden (implemented by PARC)

@@

```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'TEXT')

@@

```{r supTab_CTmislabels, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}

@@

#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(predictions %>%  
           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
           select(-cell_type_id) %>% 
           rename(PredCellType = CellType) %>% 
           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
           filter(!is.na(TrueCellType)) %>% 
           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
           group_by(PredCellType, TrueCellType) %>% 
           summarise(Count = n()) %>% 
           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
           filter(Ratio > 0.05) %>% 
           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
           flextable() %>% 
           width(width = 2), 
         size = 8, 
         part = 'all')
```
